---
title: 'Step2: Calculate Expression Similarities'
author: "Yiqun Wang"
output: github_document
---
In this example, we will 
Calculate the expression similarities between enriched genes identified from step 1 in the zebrafish notochord and hatching gland. Similarities will be calculated with several different metrics and stored in a MIMIR object.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
source("../functions/expression_similarity_calculations.R")
source("../functions/MIMIR_obj.R")
```

## Read in expression matrix for notochord and hatching gland
The expression matrices were extracted from the URD reconstructed transcriptional trajectories for hatching gland and notochord from 3 to 24 hpf. The trajectory data (URD object DraftTree.rds) can be found in the example_data folder. Cell names in the expression matrices are replaced with their pseudotimes. 
```{r read}
## read in read normalized log scaled expression for notochord (noto) and hatching gland (pcp, prechordal plate)
noto.exp = read.csv("../example_data/notochord_GeneByPseudotime.csv",as.is = T, stringsAsFactors = F,check.names = F,row.names = 1)
pcp.exp = read.csv("../example_data/hatching_gland_GeneByPseudotime.csv",as.is = T, stringsAsFactors = F,check.names = F,row.names = 1)
```

## Create an object to store expression data and similarities
```{r}
noto.obj <- createMIMIR(name="Notochord", exp.data=as.matrix(noto.exp))
pcp.obj <- createMIMIR(name="Hatching Gland", exp.data=as.matrix(pcp.exp))

noto.obj
pcp.obj
```

## Calculate expression distances using a few different methods
```{r}
## results will be stored in 3D arrarys (gene x gene x method)
### Hatching gland
pcp.obj@exp.dis <- exp_dis(x=pcp.obj@exp.data, methods=c("cosine", "soft_cosine", "euclidean", "JS"), 
                       smooth=F, pt=pcp.obj@pseudotime, sd = 0.01, cutoff=4,
                       plot_blur = T, verbose = T)

### Notochord
noto.obj@exp.dis = exp_dis(x=noto.obj@exp.data, methods=c("cosine", "soft_cosine", "euclidean", "JS"), 
                       smooth=F, pt=noto.obj@pseudotime, sd = 0.01, cutoff=4,
                       plot_blur = T, verbose = T)
```

### inspect the structure of the output
```{r}
pcp.obj@exp.dis[1:5,1:5,1:3]
```

## Convert expression distances to similarities
```{r}
## Notochord
noto.obj@exp.sim=all_dist_to_sim(noto.obj@exp.dis, max.score = 0.95)

## Hatching gland
pcp.obj@exp.sim=all_dist_to_sim(pcp.obj@exp.dis, max.score = 0.95)

## inspect  output
pcp.obj@exp.sim[1:5,1:5,1:3]
range(pcp.obj@exp.sim)
```

## Save the MIMIR objects with the expression matrices and gene expression similarities
```{r}
saveRDS(noto.obj, "../example_results/noto_mimir.rds")
saveRDS(pcp.obj, "../example_results/hg_mimir.rds")
```
