---
title: 'Step5: Check clustering results'
author: "Yiqun Wang"
output:
  github_document:
    number_sections: true
always_allow_html: true
---

In this example, we will examine several metrics of the clustering results obtained from Step4, such as number of clusters, number of member genes in clusters, and the agreement between clustering results and manual curation for a subset of notochord genes.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE}
source("../functions/MIMIR_obj.R")
source("../functions/cluster_evaluation.R")
```

# Load previous results
```{r}
noto.obj <- readRDS("../example_results/noto_mimir.rds")
```

# Load manually curated notochord enriched gene subset (300 out of 806 genes)
Genes in this test subset were grouped into modules manually by inspecting their gene expression, functional annotations, and literature. 11 genes were not assigned to any module due to a lack of functional information and no strong expression similarity with other annotated genes. 289 genes were assigned to 55 modules in this test.
```{r}
module_subset=read.csv("../example_data/curated_subset300.csv",stringsAsFactors = F)
head(module_subset)
```

## Calculate some quality metrics for each clustering result
For each clustering result, we will calculate 1. number of clusters, 2. number of singlets, 3. number of genes in the largets cluster, 4. percent enriched genes in non-singlet clusters, as well as the AMI score which measures the agreement between the clustering result and the manually curated modules for the 289 test gene subset. AMI is a metric of partition agreement, which returns a value of 1 when two partitions are identical, while random partitions score 0 on average.

### Annotations from additional databases can be used for cross validation
Here we use annotation tables obtained from MSigDB (https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp) for cross checking the results
```{r}
msigdb_c2 <- read.table("../example_data/msigdbr_C2_CGP.txt", header = 1, stringsAsFactors = F,sep = "\t") #Gene sets represent expression signatures of genetic and chemical perturbations (human data). 

#reduce to include only the genes in the embryo (union of notochord and background genes)
pcp.genes=scan("../example_results/genes.pcp.txt",what="character") #genes enriched in the hatching gland (calculated in step1). 
bg.genes=scan("../example_data/genes.bg.2x.txt", what="character") #union of genes enriched in all other cell types
msigdb_c2=msigdb_c2[which(msigdb_c2$gene%in%union(noto.obj@genes, bg.genes)),]
head(msigdb_c2)
```
### Calculate some stats for each clustering result
The calculated stats and the clustering parameters will be stored in a dataframe (returned as a data.frame if `return_df_only=T`, or as a slot in the resulting mimir object if `return_df_only=F`). 
```{r, eval=F}
noto.obj=assess_result(noto.obj, manual_modules=module_subset, return_df_only=F, 
                       internal_db_use="GO", go_cat=c("BP","MF"), external_db_tbl=msigdb_c2) # calculating annotation enrichment can be slow if the gene annotation table used as internal_db_use or external_db_tbl are large (due to a large number of annotation terms to test).
```

### Save the object with the cluster stats added
```{r, eval=F}
saveRDS(noto.obj, "../example_results/noto_mimir.rds")
```

## Plot the metrics
```{r}
noto.obj <- readRDS("../example_results/noto_mimir.rds")
head(noto.obj@cluster_metrics)
```
### Plot the number of clusters and size of max cluster
```{r}
plotly_scatter(noto.obj@cluster_metrics, x="n_cluster", y="max_cluster",color="similarity_mode", hover_text = "similarity", 
               addline_x = 150, addline_y = 150, figsize = c(500,400)) 
# good results should have a reasonable number of clusters (<150) and without most genes ending up in a single cluster (max cluster size <150)
plotly_scatter(noto.obj@cluster_metrics, x="n_cluster", y="max_cluster",color="cluster_method", hover_text = "similarity",
               addline_x = 150, addline_y = 150, figsize = c(500,400))
```

### Plot percentage of enriched gene in clusters vs agreement with manual curation (AMI)
Only keep the clustering results with reasonable # clusters (<150) and size of the largest cluster (<150 genes).
```{r}
metric_tbl=noto.obj@cluster_metrics
metric_tbl=metric_tbl[metric_tbl$n_cluster<150,]
metric_tbl=metric_tbl[metric_tbl$max_cluster<150,]

plotly_scatter(metric_tbl, x="pct_gene_in_cluster", y="AMI",color="similarity_mode", hover_text = "similarity", 
               addline_x = 90, addline_y = 0.6, figsize = c(500,350)) 

plotly_scatter(metric_tbl, x="pct_gene_in_cluster", y="AMI",color="cluster_method", hover_text = "similarity", 
               addline_x = 90, addline_y = 0.6, figsize = c(500,350))
```

### Pick a clustering result as the starting point for gene module curation
Only consider the results with >90% genes in non-singlet clusters. Pick the result with the highest agreement with the gene subset manual curation.
```{r}
metric_use=metric_tbl[metric_tbl$pct_gene_in_cluster>90,]
metric_use=metric_use[order(metric_use[["AMI"]], decreasing = T),][1,]
metric_use
```

### Store this clustering result in the `@gene.info` slot in the MIMIR object for convinient access
```{r}
noto.obj<-add_cluster_result(noto.obj, metric_use)
head(noto.obj@gene.info)
```

### Other metrics and plots can be checked to assist the decision on which similarity scores and clustering parameters to use
For instance, expression coherence within each cluster can be measured by the mean correlations between genes in the cluster. Functional coherence can be checked with additional functional annotation data that wasn't used in our previous functional similarity calculations. 

#### Cross check with the C2 gene annotations from MSigDB 
For details on MSigDB C2, see https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp. These annotations represent expression signatures of genes in genetic and chemical perturbations. We used enrichment test to identify the most significantly enriched MSigDB C2 annotations in each MIMIR cluster, and took the mean -log(p-value) of the most enriched annotations across clusters to estimate how MIMIR clusters represent functional gene sets. 
```{r}
metric_tbl=noto.obj@cluster_metrics
metric_tbl=metric_tbl[metric_tbl$n_cluster<150,]
metric_tbl=metric_tbl[metric_tbl$max_cluster<150,]
metric_tbl=metric_tbl[metric_tbl$pct_gene_in_cluster>90,]
bestxy=c(metric_use[,'msigdb_c2_median(-logP)'], y=metric_use[,'intra_mean_exp_cor'])
fig <- plotly_scatter(metric_tbl, x="msigdb_c2_median(-logP)", y="intra_mean_exp_cor",color="AMI", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 12, symbol = 'circle-open', color="coral"), showlegend = FALSE)
plotly_scatter(metric_tbl, x="msigdb_c2_median(-logP)", y="intra_mean_exp_cor",color="similarity_mode", hover_text = "similarity", 
                      marker_size = 7, figsize = c(500,350))
fig <- plotly_scatter(metric_tbl, x="msigdb_c2_median(-logP)", y="intra_mean_exp_cor",color="n_cluster", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 12, symbol = 'circle-open', color="coral"), showlegend = FALSE)
```

*The previously picked 'best' clustering result is circled (red) on the plot.*
*Some clustering results with high MSigDB C2 term enrichment (upper right points on plot) exhibit low agreement with manually curated gene subsets, as well as low cluster numbers. It is likely that these clustering results contained large clusters that combined genes from several functions.*

#### Check with C3 annotation from MSigDB
MSigDB C3 annotate genes based on shared GTRD predicted transcription factor binding sites in the region -1000,+100 bp around the TSS.
```{r, eval=F}
msigdb_c3 <- read.table("../example_data/msigdbr_C3_TFT-GTRD.txt", header = 1, stringsAsFactors = F,sep = "\t") 
msigdb_c3=msigdb_c3[which(msigdb_c3$gene%in%union(noto.obj@genes, bg.genes)),]
head(msigdb_c3)
noto.obj <- add_external_assess(noto.obj, return_df_only=F, external_db_tbl=msigdb_c3) 
saveRDS(noto.obj, "../example_results/noto_mimir.rds")
```
```{r, eval=F}
noto.obj <- readRDS("../example_results/noto_mimir.rds")
```

```{r}
metric_tbl=noto.obj@cluster_metrics
metric_tbl=metric_tbl[metric_tbl$n_cluster<150,]
metric_tbl=metric_tbl[metric_tbl$max_cluster<150,]
metric_tbl=metric_tbl[metric_tbl$pct_gene_in_cluster>90,]
metric_use=metric_tbl[metric_tbl$pct_gene_in_cluster>90,]
metric_use=metric_use[order(metric_use[["AMI"]], decreasing = T),][1,]
bestxy=c(metric_use[,'msigdb_c3_median(-logP)'], y=metric_use[,'intra_mean_exp_cor'])
fig <- plotly_scatter(metric_tbl, x="msigdb_c3_median(-logP)", y="intra_mean_exp_cor",color="AMI", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 12, symbol = 'circle-open', color="coral"), showlegend = FALSE)
plotly_scatter(metric_tbl, x="msigdb_c3_median(-logP)", y="intra_mean_exp_cor",color="similarity_mode", hover_text = "similarity", 
                      marker_size = 7, figsize = c(500,350))
fig <- plotly_scatter(metric_tbl, x="msigdb_c3_median(-logP)", y="intra_mean_exp_cor",color="n_cluster", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 12, symbol = 'circle-open', color="coral"), showlegend = FALSE)
```

### Inspect gene expression and functional annotations for each cluster
#### Use the previously picked 'best' clustering result (highest AMI)
```{r}
## Compute a smoothed version of gene expression for visualization (optional)
source("../functions/expression_similarity_calculations.R")
noto.obj@smoothed.exp=gauss.blur.matrix(noto.obj@exp.data, pt=noto.obj@pseudotime, sd=0.01, return_nn = F) # this can be automatically calculated in the plotClusters function when exp_use="smoothed.exp"
## Use the previously picked 'best' clustering result (stored in the gene.info slot)
cluster_plots = plotClusters(noto.obj, cluster_use="STRING_ExpTxt+GO_noCC+Reactome+Interpro:AND:JS_sim;leiden_6_RBConfigurationVertex",
                             exp_use="smoothed.exp", save_pdf="../example_results/cluster_plots1.pdf", 
                             save_width = 20, save_height = 40, return_plots=T, show_plot=F) ## show_plot=F: full plot is too big to show at once in this notebook
```

*Full plot saved to `../example_results/cluster_plots1.pdf`*

**Visualize cluster plots**
```{r, fig.width = 20, fig.height = 24}
library(patchwork)
wrap_plots(cluster_plots[1:20], ncol = 4)
wrap_plots(cluster_plots[21:40], ncol = 4)
wrap_plots(cluster_plots[41:length(cluster_plots)], ncol = 4)
```

*This clustering strategy resulted in 54 non-singlet clusters. For each cluster, traces in the left graph represent expression of each gene in the cluster. Expression levels are not scaled in this plot, (a scaled version might show the result better since JS and cosine similarity calculation is not sensitive to the absolute expression levels). The top 3 most enriched functional annotation terms in a cluster were identified and the number of annotated genes are shown in the Venn diagrams on the right. The annotation terms tested are from GO, Reactome, and Interpro (using data tables stored in `noto.obj@annotations`, saved in step3). Genes in a cluster that are not annotated by the top 3 annotations are indicated by the number outside the circles (upper right corner in the box). In most cases, we observe that __1.__ the 3 top annotations are related or redundant, indicating that they represent the same or related biological functions; and __2__ the majority of genes in a cluster are covered by the top 3 annotations, indicating that most genes in a cluster are from the same or related biological processes.*

#### Use the clustering result with high enrichment of MSigDB C2 annotations
(the red circled point in the following plot). MSigDB C2 annotations are based on expression signatures of genes in genetic and chemical perturbations.
```{r}
fig <- plotly_scatter(metric_tbl, x="msigdb_c2_median(-logP)", y="intra_mean_exp_cor",color="AMI", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
best_metric=metric_tbl[intersect(which(metric_tbl$intra_mean_exp_cor>0.14),which(metric_tbl$`msigdb_c2_median(-logP)`>7)),]
bestxy=c(best_metric$`msigdb_c2_median(-logP)`,best_metric$intra_mean_exp_cor)
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 15, symbol = 'circle-open', color="coral"), showlegend = FALSE)
```
**Extract this clustering result and write it to `noto.obj@gene.info`**
```{r}
noto.obj<-add_cluster_result(noto.obj, best_metric)
head(noto.obj@gene.info)
```

**Generate cluster plot**
```{r, fig.width = 16, fig.height = 16}
cluster_use="GO+Reactome+Interpro:OR:soft_cosine_sim;leiden_2_RBConfigurationVertex"
cluster_plots = plotClusters(noto.obj, cluster_use=cluster_use, exp_use="smoothed.exp", save_pdf=F, 
                             save_width = 20, save_height = 40, return_plots=T, show_plot=F) 

wrap_plots(cluster_plots[1:length(cluster_plots)], ncol = 3)
```

*This clustering strategy resulted in only 9 clusters. Each cluster is big. Gene expression seems relatively consistent withing cluster and different between clusters. While the top 3 enriched functional annotation terms are highly related in most clusters, they only cover a small portion of all genes in each cluster (genes in cluster but not annotated by the top 3 annotations are indicated by the number outside the circles in the Venn diagram). This indicates that each cluster contains genes from diverse biological processes.*

#### Use the clustering result with high enrichment of MSigDB C3 annotations
(the red circled point in the following plot). MSigDB C3 annotate genes based on shared GTRD predicted transcription factor binding sites in the region -1000,+100 bp around the TSS.
```{r}
best_metric=metric_tbl[intersect(which(metric_tbl$intra_mean_exp_cor>0.11),which(metric_tbl$`msigdb_c3_median(-logP)`>1.7)),]
bestxy=c(best_metric$`msigdb_c3_median(-logP)`,best_metric$intra_mean_exp_cor)

fig <- plotly_scatter(metric_tbl, x="msigdb_c3_median(-logP)", y="intra_mean_exp_cor",color="AMI", hover_text = "similarity", 
               color_pal = 'viridis', marker_size = 7, figsize = c(500,350))
fig %>% add_trace(type = 'scatter', mode = 'markers', x = bestxy[1], y = bestxy[2],color='',
                  marker = list(size = 12, symbol = 'circle-open', color="coral"), showlegend = FALSE)
```

**Extract this clustering result and write it to `noto.obj@gene.info`**
```{r}
noto.obj<-add_cluster_result(noto.obj, best_metric)
head(noto.obj@gene.info)
```

**Generate cluster plot**
```{r, fig.width = 20, fig.height = 20}
cluster_use="STRING_ExpTxt+GO_noCC+Reactome+Interpro:+:soft_cosine_sim;leiden_6_RBConfigurationVertex"
cluster_plots = plotClusters(noto.obj, cluster_use=cluster_use, exp_use="smoothed.exp", save_pdf=F, 
                             save_width = 20, save_height = 40, return_plots=T, show_plot=F) 

wrap_plots(cluster_plots[1:16], ncol = 4)
wrap_plots(cluster_plots[17:32], ncol = 4)
wrap_plots(cluster_plots[32:length(cluster_plots)], ncol = 4)
```

*The "best" clustering result picked based on intra-cluster gene correlation and MSigDB C3 annotation enrichment is similar to that picked based on the highest agreement with our manually curated gene subset. This shows that in absence of curated gold-standard, expression and functional coherence measured with additional approaches/data sources (that are not used in calculating the integrated similarity for clustering) can be used to pick the best clusters for further manual curation.*

### Save the best clustering result (based on agreement with manually curated gene subset)
```{r}
cluster_use="STRING_ExpTxt+GO_noCC+Reactome+Interpro:AND:JS_sim;leiden_6_RBConfigurationVertex"
auto_module = get_auto_module(noto.obj,cluster_use=cluster_use, name_by_enriched=T, use_description=T)
module2tsv(auto_module, "../example_results/auto_module.tsv")

## save as an rds (can be loaded into the shiny app for manual curation)
saveRDS(auto_module,"../example_results//Cluster_basis_for_Curation.rds")
```

